---
title: "Focusing your audiences attention"
author: "Gareth Green"
output: slidy_presentation
---

```{r echo = FALSE}
# Course: 5210 Communicating Data
# Purpose: SWD Ch 4 Focusing your audience's attention
# Date: August 5, 2019
# Author: Gareth Green

```

```{r echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r message = FALSE, echo = FALSE}
# Load packages
library(tidyverse) 

```

```{r echo = FALSE}
# Load data
dis <- read.csv("Disability.csv")

# Recode date to age range, this is to fix an Excel formatting problem
dis <- dis %>% 
  mutate(age_group = recode(age_group, `12-Jun` = "6-12"))

```


Storytelling with Data Ch 4: Focusing your audience's attention
===============================================

<div style="float: right; width: 95%;">

+ EDA and modeling are about finding the important message

    - but that is not enough
    - you need to invest sufficient effort to communicate the message/finding

+ We have discussed and demonstrated:

    1. best practices for communication and designing visuals
    2. how to choose the appropriate visual for your data and message
    3. how to reduce clutter so your message is clear

+ Now we will detail how to focus your audiences attention 

  - make sure they understand the message/finding you are trying to communicate
  - make it easy to understand your message/finding

</div>


Focus your audience's attention
===============================================

<div style="float: right; width: 95%;">

+ You are working on your presentation visuals

    - you have considered your audience and the relevant context
        - i.e. what is the goal, target, or comparison
    - you have determined an effective visual for your data and message
        - recall the basic rules of bar graphs, scatterplots, line graphs, etc...
    - you have removed the excess clutter from the visual
        - the stuff that R, Excel, PowerBI, Tableau add for no reason

+ Now, how do you make sure the audience sees what you want them to see?

    - it is time to consider what your audience sees and remembers
    
  >Preattentive Attributes: size/shape, color, and position

    1. Use these to direct your audiences attention
    2. Use to build a visual hierarchy to lead your audience through the information

</div>


Focus your audience's attention  
===============================================

<div style="float: right; width: 95%;">

+ Preattentive attributes signal where to look

    - consider the following exmample from SWD

How many 3's are in the block of numbers?

> 7 5 6 3 9 5 0 6 8 4 7 3  
> 6 5 8 6 6 3 0 3 7 5 7 6  
> 8 6 0 3 7 2 6 5 8 6 0 2  
> 8 4 6 5 8 9 1 0 7 8 3 0  

+ Its not easy to count because they blend in

+ Now we will increase the intensity of the **3**s
    
</div>


Focus your audience's attention  
===============================================

<div style="float: right; width: 95%;">

How many 3's are in the block of numbers?

> 7 5 6 **3** 9 5 0 6 8 4 7 **3**  
> 6 5 8 6 6 **3** 0 **3** 7 5 7 6  
> 8 6 0 **3** 7 2 6 5 8 6 0 2  
> 8 4 6 5 8 9 1 0 7 8 **3** 0  

+ Your eyes jump from one 3 to the next

    - think about how to the same thing with your message

<br>

>**Don't just show the data, use the data to tell a story**  

</div>


Focus your audience's attention - preattentive attributes
===============================================

<div style="float: right; width: 95%;">

+ Using **titles, color (contrast), size/shape, and annotations**

+ Your audience will see these elements first

    - you can create a visual hierarchy
    - must use sparingly, otherwise it just looks cluttered
      - produce cognitive overload

+ Using preattentive attributes are good in explanatory visuals

    - do not use in EDA
    - may *hide* something you should discover

In text color or bold are generally more effective

  - italics don't really show up
  - underline and text boxes look cluttered

</div>


Focus your audience's attention - titles  
===============================================

<div style="float: right; width: 95%;">

+ We've talked about how experts read individual visuals

    - title, axes labels, and legend

+ Consider this with the positions on your visual - the title is critical

  - **don't** just have "Disability Analysis" or "Expenditures by age" as a title
  - tell why the visual is important
      - why your audience should care

+ Highlight main points

  - let sub-points fade into the background
  - surprising how software like PowerBI does the opposite

+ Similar on a dashboard 

  - we will discuss focusing your audiences attention with respect to dashboards in week 7

</div>


Focus your audience's attention - color  
===============================================

<div style="float: left; width: 50%;">

+ Color in graphs is one of the strongest tools

    - emphasize the important elements of your visual
        - the elements that support our message
    - consider deemphasizing less important elements 
        - keep for less important for context
        - or elimiate less important if you have good reason

</div>

<div style="float: left; width: 50%;">

+ Consider strategic use of color

    - use gray as a base color rather than black
    - use a single color palatte
      - create contrast within a single palatte rather than stark color contrast
      - do not use red and blue together unless you want to show divergence
    - use color sparingly
    - use color consistently
    - consider people that are colorblind
         - hard to see shades of red and green together
    - blues or individual color schemes with different shades

</div>


Focus your audience's attention - size/shape 
===============================================

<div style="float: left; width: 50%;">

+ We often use size/shape without realizing it

  - the height of bar in graphs and histograms is size
  - size/shape of points in a scatter plot
  - line thickness can be used to emphasize a particular line
    
+ I do not use size/shape in individual visuals much

  - the eye does not detect small differences in size/shape
  - sometimes using different size/shapes looks like clutter and is distracting

</div>

<div style="float: left; width: 50%;">

+ Often use size to highlight text you want to standout

  - for example, titles are usually larger than other text  
  - though color or intensity (bold) can be more effective

+ Size can be important in dashboards

    - the eye is drawn to a larger graph
    - use your dash layout to highlight your main point
    - use smaller visuals to provide context

</div>


Focus your audience's attention - annotations
=========================================

<div style="float: left; width: 45%;">

+ Annotations are methods to make portions of the data stand out

  - typical examples are text and shading
  
+ Text can be used within a plot to label specific data points

  - must be careful to not introduce clutter
  - usually only want to label one or several data points
  - label regions of a graph

</div>

<div style="float: left; width: 55%;">

+ Shading or enclosing can be used to highlight regions of a graph

  - indicate a specific attribute in the data
    - uncertainty like a standard error ribbon
    - sub-group of data important to your analysis 
  - use to create enclosure or indicate similarity
  - indicate different periods of time
    - separate observed vs forecast data

+ Text and shading together

  - shading to make the data stand out
  - text to explain what it is or why it is important

</div>


But "_how do I do it in R??_"
===============================================

<div style="float: right; width: 95%;">

+ You may have ideas of how to use **titles, color (contrast), size/shape, and annotations**

  - but how actually do it??
  - it is one thing to know what you want to do and another to do it software
  
+ This is a problem in all software

  - I have spent countless hours in Excel trying to make tables and graphs look just right
  - one of the benefits of R is that you can set up themes and functions to do these things for you
  - once you determine a preferred format, make your own library of code to create preferred visual effects
  
+ We will spend the rest of today working on methods to focus your audiences attention

  - but remember, there is more than one way to accomplish something so you may find another way
  - use whichever method you feel is most effective
  - every time I teach a class I learn a new method that a student discovers
    - packages are updated and new ones created

</div>


