---
title: "Shape, size and linetype"
author: "Gareth Green"
output: slidy_presentation
---

```{r echo = FALSE}
# Course: 5210 Communicating Data
# Purpose: Demonstrate the use of Shape, size and linetype
# Date: August 5, 2019
# Author: Gareth Green

```

```{r echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r message = FALSE, echo = FALSE}
# Load packages
library(tidyverse) 

```

```{r echo = FALSE}
# Load data
dis <- read.csv("Disability.csv")

# Recode date to age range, this is to fix an Excel formatting problem
dis <- dis %>% 
  mutate(age_group = recode(age_group, `12-Jun` = "6-12"))

# Remove small sample ethnicities
dis <- dis %>% 
  filter(ethnicity %in% c("Asian", "Black", "Hispanic", "White"))

```


Shape, size and linetype
=========================================

<div style="float: right; width: 95%;">

+ Size and shape are most commonly used in scatter plots

  - size and shape can be difficult to detect
  - points tend to blend together and overlap

+ Linetype are most commonly used in line graphs

  - linetypes are good for several variables 
  - too many creates clutter
  - often linetypes can look quite similar

+ Size can be used on quantiative variables

+ Shape and linetype can be used on factor variables

  - different shape/linetype for each factor level
  - can sometimes use thickness to make a specific factor level stand out

+ I recommend using size and shape sparingly

  - linetypes are good for 5 or fewer factor levels

</div>


Shape
=========================================

<div style="float: left; width: 55%;">

+ Shape is set with the `shape =` attribute in the `aes()` of the `ggplot()` or `geom_point()` functions

+ This scatter plot helps illustrate when to use shape

  - below age 25 there are so many overlapping points that we cannot see the different shapes
    - don't use shape when there are many overlapping points
  - above age 25 the points spread out
    - okay to use shape when points are spread out
    
+ However, it can still be difficult to see shapes when spread out

  - can you easily see the difference between squares and circles?

```{r shape, eval = FALSE}
# Basic scatter plot with base shape points
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity)) +
  geom_point() +
  theme_classic()

```

</div>

<div style="float: left; width: 45%;">

```{r ref.label = "shape", echo = FALSE}
# Basic scatter plot with different shape points

```

</div>


Variety of shapes
=========================================

<div style="float: left; width: 50%;">

+ There are a variety of shapes available in ggplot

  - below is a list by shape number
  
```{r echo = FALSE}
# Show a list of available shapes
data.frame(shape = 0:24) %>% 
  ggplot(mapping = aes(0, 0, shape = shape)) +
  geom_point(aes(shape = shape), size = 5, fill = "red") +
  scale_shape_identity() +
  facet_wrap( ~ shape) +
  theme_void()

```

+ Can set custom shapes

</div>

<div style="float: left; width: 50%;">

```{r}
# Basic scatter plot with different shape points and base size
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity)) +
  geom_point() +
  theme_classic() +
  scale_shape_manual(values = c(0, 4, 16, 3)) # set custom points

```

</div>


Size
=========================================

+ There are two ways to use size - set base size or distinguish different data

<div style="float: left; width: 50%;">

+ To set the base size of points set `size = #`

```{r}
# Scatter plot with different shapes but larger size
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity)) +
  geom_point(size = 3) +
  theme_classic()

```

</div>

<div style="float: left; width: 50%;">

+ To distinguish different data set size in the `aes()` function

```{r}
# Set size by magnitude of data
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, size = expenditures)) +
  geom_point() +
  theme_classic()

```

</div>


Shape and size
=========================================

<div style="float: left; width: 50%;">

+ You can set size by a factor variable as well, but it is not advised

  - note size is set within the `aes()` function because you are mapping the aesthetic to a characteristic of the data

+ Note, the multiple legends clutter the graph

```{r shape_size1, eval = FALSE, warning = FALSE}
# Basic scatter plot with different size and shape points
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity, size = factor(ethnicity))) +
  geom_point() +
  theme_classic() 

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "shape_size1", echo = FALSE, warning = FALSE}
# Call previous graph

```

</div>


Custom shape and size
=========================================

<div style="float: left; width: 50%;">

+ You can set shape and size and combine into one legend

  - using the `scale_shape_manual()` and `scale_size_manual()` creates custom shapes and sizes
    - setting `values = c()` gives the custom shape and size
    - including "" combines the legends into one

```{r shape_size2, eval = FALSE}
# Basic scatter plot with different shape points
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity, size = factor(ethnicity))) +
  geom_point() +
  theme_classic() +
  scale_size_manual("", values = c(4, 2, 3, 1)) +
  scale_shape_manual("", values = c(0, 4, 16, 3))

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "shape_size2", echo = FALSE, message = FALSE}
# Call previous graph

```

</div>


Shapes with color and size
=========================================

<div style="float: left; width: 50%;">

```{r}
# Basic scatter plot with different shape and color points
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity, color = ethnicity)) +
  geom_point() +
  theme_classic()

```

+ Using color to further distinguish points

  - recall, you must set `color = ethnicity` within `aes()`
  - though it is easier to see the difference, this is due to color not shape

</div>

<div style="float: left; width: 50%;">

```{r}
# Basic scatter plot with base shape and larger size points
dis %>%   
  ggplot(mapping = aes(x = age, y = expenditures, shape = ethnicity)) +
  geom_point(size = 2) +
  theme_classic()

```

+ Here we are increasing the size of all points so it is easier to see them

  - not distinguishing different data by size
  - set size outside of the `aes()` function

+ This is good and bad - can see the shape but more overlap
  
</div>


Linetype
=========================================

<div style="float: left; width: 50%;">

+ Line graphs are best used to illustrate continuity

  - most commonly used in time series data
  - we will add a date variable to illustrate the use of linetypes
  
```{r}
# Create random dates for making line graphs
dis$date <- sample(seq(as.Date("2015/01/01"), as.Date("2018/01/01"), by = "month"), nrow(dis), replace = TRUE)

```

+ Two things to note:

  1. a standard line graph is not good for this data - **spaghetti graph**
  2. it can be difficult to decifer the base linetypes in ggplot

```{r linetype1, eval = FALSE}
dis %>% 
  group_by(date, ethnicity) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = date, y = med_exp, linetype = ethnicity)) +
  geom_line() +
  theme_classic()

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "linetype1", echo = FALSE, message = FALSE}
# Call previous graph

```

</div>


Variety of linetypes
=========================================

<div style="float: left; width: 95%;">

+ There are a variety of linetypes available in ggplot

  - below is a list by linetype number
  
```{r echo = FALSE}
# Show a list of available linetypes
data.frame(linetype = factor(1:13)) %>% 
  ggplot(mapping = aes(0, 0, linetype = linetype)) +
  geom_segment(mapping = aes(x = 0, xend = 1, y = linetype, yend = linetype, linetype = linetype), size = 1.1)+
  scale_x_continuous(limits = c(0, 1), breaks = NULL, name = "") +
  scale_y_discrete(name = "") +
  guides(linetype = guide_legend(reverse = TRUE)) +
  theme_void()

```


</div>


Improving readability with linetype
=========================================

<div style="float: left; width: 50%;">

+ Set custom linetype and size manually

```{r}
# Custom linetypes
dis %>% 
  group_by(date, ethnicity) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = date, y = med_exp, linetype = ethnicity, size = ethnicity)) +
  geom_line() +
  theme_classic() +
  scale_linetype_manual(values = c(1, 3, 6, 10)) +
  scale_size_manual(values = c(0.5, 1, 1.25, 1))

```

</div>

<div style="float: left; width: 50%;">

+ Seperate line graphs using facets

```{r}
# Using facets
dis %>% 
  group_by(date, ethnicity) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = date, y = med_exp, linetype = ethnicity)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ ethnicity) +
  scale_linetype_discrete(guide = FALSE)

```

</div>


Wrap up
=========================================

<div style="float: right; width: 95%;">

+ You have to be careful using shape, size and linetype

  - easy to create clutter
  - can make a graph difficult to read
  - **remember**, your goal is to make it as easy as possible for the audience to understand your main point
    - adding too much to a graph can have the opposite effect

+ Remember, just becasuse you can do something doesn't mean you should

  - only add differences in shape, size and linetype if they make it easier for your audience

</div>


