---
title: "Annotating graphs: Shading, lines and text"
author: "Gareth Green"
output: slidy_presentation
---

```{r echo = FALSE}
# Course: 5210 Communicating Data
# Purpose: Demonstrate shading, lines and text in graphs
# Date: August 5, 2019
# Author: Gareth Green

```

```{r echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r message = FALSE, echo = FALSE}
# Load packages
library(tidyverse) 

```

```{r echo = FALSE}
# Load data
dis <- read.csv("Disability.csv")

# Recode date to age range, this is to fix an Excel formatting problem
dis <- dis %>% 
  mutate(age_group = recode(age_group, `12-Jun` = "6-12"))

# Remove small sample ethnicities
dis <- dis %>% 
  filter(ethnicity %in% c("Asian", "Black", "Hispanic", "White"))

```

Annotations
=========================================

<div style="float: right; width: 95%;">

+ What is an annotation?

  - a written note or shading added to text, table or graph
  - objective is to clarify or highlight a point
  - make an element standout
  
+ We will only use the `annotate()` function

  - other functions include: `geom_text()`, `geom_polygon()`, `geom_rect()`, etc
  - additional packages: `gghighlight`, `ggforce`, `ggrepel`, `directlabels`, etc

+ Use carefully

  - adding to much text can clutter
  - shading can make elements difficult to read

+ Note, as we work through these graphs we are not doing all formatting discussed earlier to reduce code lines on the slides

</div>

Base graphic prior to annotation
=========================================

<div style="float: left; width: 50%;">

+ What are several ways we can improve this graph to make it easier for our audience to understand our main point?

```{r ann1}
# The base graph we have been using
g_ann <- dis %>% 
  group_by(ethnicity, gender) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = reorder(ethnicity, med_exp), y = med_exp, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_discrete(limits = c( "Male", "Female")) +
  theme_classic()

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "ann1", echo = FALSE}
# Retrieve previous graph
g_ann

```

</div>


Annotations as context
=========================================

<div style="float: left; width: 50%;">

+ One use of annotations is to provide context

    - context can set the visual in the larger picture 
    - add information to help with interprestation of the visual
    - here I add a line and text

+ Here we provide the overall median expenditures as context to compare median expenditures of each ethnicity

    - note that we are using hline and reversing x and y since we use `coord_flip()` prior to adding these layers
    - if we added these annotations prior to adding `coord_flip()` we would not need to reverse our annotations

+ Location of the text is done by setting x and y

    - the height is set as x = "Hispanic", which is based on the center of that factor level
        - I could also set x = 1 to get the same location
    - note we set hjust = 1 to move the text right of center
        - I have also added 200 to the y location to move the lable off of the line

</div>

<div style="float: left; width: 50%;">

```{r}
# The base graph we have been using
g_ann + 
  geom_hline(yintercept = median(dis$expenditures)) +
  annotate(geom = "text", x = "Hispanic", 
           y = 200 + median(dis$expenditures), 
           label = "Median expenditures", hjust = 0) 

```

</div>


Shading to highlight a finding
=========================================

<div style="float: left; width: 50%;">

+ Use a shaded shape to highlight an area on the graph

    - suppose we want to highlight that white and black male have higher median expenditures 
    - we can add two custom shaded rectangles
    - we could then add text or mention in the title/sub-title
    
+ I have set the y coordinates of the rectangle based on calculations

    - could also just set hard numbers, but if the data changed I'd have to adjust

+ x-coordinates are set with location of factor levels in the visual

    - can you find a way to base it off of the actual factor level?

+ `fill =` sets the color of the shading

    - `color =` sets the color of the outline of the shaded shape
    - `alpha = 0.5` sets the transparancey of the shading

</div>

<div style="float: left; width: 50%;">

```{r }
# The base graph we have been using
g_ann + 
  annotate(geom = "rect", xmin = 3.5, xmax = 4.5,
           ymin = -200, ymax = 200 + median(dis$expenditures[dis$ethnicity == "White" & dis$gender == "Male"]),
           fill = "lightblue", alpha = 0.5) +
  annotate(geom = "rect", xmin = 1.5, xmax = 2.5,
           ymin = -200, ymax = 200 + median(dis$expenditures[dis$ethnicity == "Black" & dis$gender == "Male"]),
           fill = "lightblue", alpha = 0.5) 

```


</div>


Shading of time-series
=========================================

```{r echo = FALSE}
# Create random dates for making line graphs
set.seed(123)
dis$date <- sample(seq(as.Date("2015/01/01"), as.Date("2018/01/01"), by = "month"), nrow(dis), replace = TRUE)

```

<div style="float: left; width: 50%;">

+ An excellent use of shading is to highlight a specific time period on a line graph

    - it tels your audience what period is most important to your point

```{r ann4, eval = FALSE}
# Look at median expenditures over time 
dis %>% 
  group_by(date) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = date, y = med_exp)) + 
  geom_line() +
  annotate(geom = "rect", xmin = as.Date("2017-09-01"), 
           xmax = as.Date("2017-11-01"), 
           ymin = 3000, ymax = 33000, 
           fill = "lightblue", alpha = 0.5) +
  theme_classic()

```

+ One thing to note is that your x-axis settings need to be converted to dates using `as.Date()`

    - in general your coordinates must be in the units of the data on the graph
    - you can add or subract "units" to shift annotations left or right

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "ann4", echo = FALSE}
# Call graph from previous chunk

```

</div>
    
    
Wrap up
=========================================

+ Annotations can be used to:

    - provide context: set the finding within the bigger picture
    - highlight findings: make it easier for your audienct to see and understand your point

+ Be careful not to add clutter to your visual

+ We have only used the `annotation()` function, you should _experiment_ with:

    - other functions include: `geom_text()`, `geom_polygon()`, `geom_rect()`, etc
    - additional packages: `gghighlight`, `ggforce`, `ggrepel`, `directlabels`, etc

+ There is a lot more you can do!!
