---
title: "Titles, labels and legends"
author: "Gareth Green"
output: slidy_presentation
---

```{r echo = FALSE}
# Course: 5210 Communicating Data
# Purpose: Demonstration of visual design step-by-step
# Date: August 5, 2019
# Author: Gareth Green

```

```{r echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r message = FALSE, echo = FALSE}
# Load packages
library(tidyverse) 

```

```{r echo = FALSE}
# Load data
dis <- read.csv("Disability.csv")

# Recode date to age range, this is to fix an Excel formatting problem
dis <- dis %>% 
  mutate(age_group = recode(age_group, `12-Jun` = "6-12"))

```


Titles, labels and legends
=========================================

<div style="float: left; width: 50%;">

+ Titles and labels are informative text used to describe the visual

  - often used in an uninformative manner
  - for example, the main title often repeats the axes
    - i.e. Median expenditures by ethnicity
    - which is informative, but repetitive so becomes clutter

+ Think of the set of labels on a visual as a whole rather than individually

  - if axes are well labeled you can use the title and subtitle to inform what the goal of the in visual

+ We do two things in this set of slides

    1. Discuss _best practices_ of titles, labels and legends
    2. How to accomplish the best practices in ggplot
      - this is a challenge regardless of the software you use

</div>

<div style="float: left; width: 50%;">

**Best practices**

+ Title/subtitle: indicate the main/sub points you want the audience to see in the visual

  - always set while graphing
  - avoid long titles, be concise

+ Axes and legend labels: variable and units (if applicable)

  - often taken from the data, but you can format factor level names or digits while wrangling or graphing
  
+ Tick and legend labels: factor level or quantity

  - format appropriately (no scientific notation, %, $, ",", etc)
  - often taken directly from the data, but you can format factor level names or digits in the data or while wrangling

+ Reference labels: additional labels that could include a figure number or source

</div>


Primary titles and labels
=========================================

<div style="float: left; width: 50%;">

+ Recall the base graph we have been redesigning

+ Can set title and change axes and legend labels with the `labs()` function

  - of course there are other ways as well: `ggtitle()`, `xlab()`, `ylab()`, change variable name in data, etc

+ Have added or updated each of these with one function

```{r title1, eval = FALSE}
# The base graph we have been redesigning plus labs() function
dis %>% 
  group_by(ethnicity, gender) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = reorder(ethnicity, med_exp), y = med_exp, fill = reorder(gender, med_exp))) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_discrete(limits = c("Female", "Male")) +
  theme_classic() +
  labs(title = "Main title\nmain title continued", 
       subtitle = "Subtitle", fill = "Legend\ntitle", 
       x = "New x-axis label", y = "New y-axis label", 
       tag = "Tag", caption = "Caption")

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "title1", echo = FALSE}
# Get the previous graph

```

</div>


A couple of things to notice
=========================================

<div style="float: left; width: 50%;">

```{r ref.label = "title1", echo = FALSE}
# Get the previous graph

```

</div>

<div style="float: left; width: 50%;">

+ Look at the additional code and graph to see which element in code relates to element in the graph

+ Additional code

      labs(title = "Main title\nmain title continued", 
           subtitle = "Subtitle", fill = "Legend\ntitle", 
           x = "New x-axis label", y = "New y-axis label", 
           tag = "Tag", caption = "Caption")
             
+ Can create a new line in text using \\n

+ Why are the x and y axis labels reversed??

+ `tag =` would likely be a figure number or something else

+ `caption = ` would likely be a source or reference

</div>


Primary titles and labels
=========================================

<div style="float: left; width: 50%;">

+ Must be careful with your titles and labels 

  - **they can clutter a visual**

+ _Figure 1_ might be better handled another way

+ I think the sub-title is too long

  - using two lines looks cluttered
  - **should remove the small sample ethnicities from the analysis**
    - could be misleading and add clutter
    - could be used to support a false narrative

+ I removed the y-axis and legend labels to reduce clutter

  - these variables are self-explanatory
  
</div>

<div style="float: left; width: 50%;">

```{r echo = FALSE}
# The base graph we have been redesigning plus labs() function
dis %>% 
  group_by(ethnicity, gender) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = reorder(ethnicity, med_exp), y = med_exp, fill = reorder(gender, med_exp))) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_discrete(limits = c("Female", "Male")) +
  theme_classic() +
  labs(title = "There is a wide variation of expenditures by ethnicity", 
       subtitle = "However, only Asian, Black, Hispanic and White have\n large enough samples to be statistically valid",
       fill = "", 
       x = "", y = "Median expenditures", 
       tag = "Figure 1", caption = "\nSource: Disability.csv data set from Week 1")

```

</div>


A remake of the overall graph
=========================================

<div style="float: left; width: 50%;">

+ Remove the small sample ethnicities from the analysis

  - problem is that Male median expenditures is much larger than Female
  - but we found this is due a difference in age profile
    - more young Female with lower expenditures and more older Males with high expenditures
  - based on our analysis, this visual is misleading 
  
+ We will continue using this visual for demonstration purposes

+ I added a note in the subtitle to _see the discussion_

  - best not to use a deceiving visual, but if you have to make sure to **debunk or discuss the deception**
  
</div>

<div style="float: right; width: 50%;">

```{r echo = FALSE}
# Name the new overall graph p so as not to have so much repetitive code
p <- dis %>% 
  filter(ethnicity %in% c("Asian", "Black", "Hispanic", "White")) %>% 
  group_by(ethnicity, gender) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = reorder(ethnicity, med_exp), y = med_exp, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_discrete(limits = c("Male", "Female")) +
  theme_classic() +
  labs(title = "There is a wide variation of expenditures by ethnicity, but not by gender", 
       subtitle = "The difference by gender for White and Black is deceiving\ndue to differences in age profiles and associated expenditures (see discussion)",
       fill = "", x = "", y = "Median expenditures", 
       caption = "\nSource: Disability.csv data set from Week 1")

p

```

</div>


Tick labels and marks
=========================================

<div style="float: left; width: 50%;">

+ We saw some tick label and mark customization earlier

```{r tick1, eval = FALSE}
# Change the axis ticks and labels using 
p <- p + # p is the overall base graph
  theme(axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_blank()) +
  scale_y_continuous(breaks = seq(0, 30000,5000))

p

```

+ Also see the following link [http://www.sthda.com/english/wiki/ggplot2-axis-ticks-a-guide-to-customize-tick-marks-and-labels](http://www.sthda.com/english/wiki/ggplot2-axis-ticks-a-guide-to-customize-tick-marks-and-labels)

  - this web site details many ways to customize your axis ticks

</div>

<div style="float: left; width: 50%;">

+ Look at the code and graph to see which element in code relates to element in the graph

```{r ref.label = "tick1", echo = FALSE}
# Get the previous graph

```

</div>


Tick labels and marks
=========================================

<div style="float: left; width: 50%;">

+ We can go one step further by loading the `scales` package

  - the `scales` package has custom formatting for tick labels
  - here we will use the `labels = dollars`

```{r tick2, message = FALSE, eval = FALSE}
# Load library
library(scales)

# Change the axis ticks and labels using 
p <- p + scale_y_continuous(breaks = seq(0, 30000,5000),
                            labels = dollar)

p

```

</div>

<div style="float: left; width: 50%;">

+ Look at the code and graph to see which element in code relates to element in the graph

```{r ref.label = "tick2", message = FALSE, echo = FALSE}
# Get the previous graph

```

</div>


Legend location and labels
=========================================

<div style="float: left; width: 50%;">

+ We've already seen how to change the legend title

  - it will usually be a variable name
  
+ Can also change the location and labels

  - note, I used `values = c("#F8766D", "#00BFC4")` to set the color

```{r legend1, message = FALSE, eval = FALSE}
# Change the legend location, title and labels 
p <- p + theme(legend.position = "bottom") +
         scale_fill_manual(labels = c("Women", "Men"), values = c("#F8766D", "#00BFC4")) 

p

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "legend1", message = FALSE, echo = FALSE}
# Get the previous graph

```

</div>


Wrap up
=========================================

<div style="float: left; width: 50%;">

+ We have only scratched the surface of what you can do for formatting

  - pretty much anything is possible if it makes sense
  - but it could take some effort to figure it out
  - there is more than one way to achieve a goal

+ There are many additional functions similar to what we have used

  - earlier and here you saw many option using the `theme()` function
  - also many `scale_xxx_yyyy()` functions
    - the `xxx` relates to an aes() attribute like x, y, fill, color, size, shape, alpha, etc...
    - the `yyy` relates to variable position along an axes like continuous, discrete, date, time, manual, etc 
    
</div>

<div style="float: left; width: 50%;">

+ For example: 

  - `scale_fill_manual(labels = c("Women", "Men"), 
                                  values = c("#F8766D", "#00BFC4"))` 
    - set custom labels and color for `fill = gender` that was specified in `aes()`
  - `scale_y_continuous(breaks = seq(0, 30000,5000),
                        labels = dollar)`
    - set custom axes breaks and labels for `y = med_exp` that was specified in `aes()`

+ You will **always** need to make changes like this to make your visuals professional and easy to read

  - each software package will require you to figure out how to make the changes
  - the good thing about _R_ is that you can use code to recreate what you have done
    - in Excel, PowerBI, Tableau, etc you can point-and-click
    - but cannot recreate the commands automatically

</div>


