---
title: "Use color wisely"
author: "Gareth Green"
output: slidy_presentation
---

```{r echo = FALSE}
# Course: 5210 Communicating Data
# Purpose: Illustrate good and bad examples of using color
# Date: August 5, 2019
# Author: Gareth Green

```

```{r echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r message = FALSE, echo = FALSE}
# Load packages
library(tidyverse) 

```

```{r echo = FALSE}
# Load data
dis <- read.csv("Disability.csv")

# Recode date to age range, this is to fix an Excel formatting problem
dis <- dis %>% 
  mutate(age_group = recode(age_group, `12-Jun` = "6-12"))

# Remove small sample ethnicities
dis <- dis %>% 
  filter(ethnicity %in% c("Asian", "Black", "Hispanic", "White"))

```


Using color in graphs
=========================================

<div style="float: left; width: 55%;">

+ Color is powerful

  - draws attention
  - shows relationships

+ Often misused

  - create a lot of clutter

+ Think of color being used four different ways

  - draw attention - make one or a set of elements stand out
  - qualitative - make no individual element stand out
  - sequential - show order of elements 
  - diverging - where elements diverge from neutral to two extremes

+ We will use the `colorspace` package to make coding easier

  - other packages include RColorBrewer, wesanderson, colorfindr and many more

</div>

<div style="float: left; width: 45%;">

```{r message = FALSE}
# Load librarys, make sure to install these packages before running the code
library(colorspace)

# Plot reset color palettes
hcl_palettes(plot = TRUE)

```

</div>


Simple version of revised base graph - counts
=========================================

<div style="float: left; width: 50%;">

+ There is no reason for color in this graph

```{r}
# Basic bar graph
dis %>% 
  ggplot(mapping = aes(x = ethnicity)) +
  geom_bar() +
  coord_flip() +
  theme_classic()

```

</div>

<div style="float: left; width: 50%;">

+ But something other than black is easier on the eyes

```{r}
# Basic bar graph with custom color
dis %>% 
  ggplot(mapping = aes(x = ethnicity)) +
  geom_bar(fill = "light blue") +
  coord_flip() +
  theme_classic()

```

</div>


Color works similarly in most ggplot geoms
=========================================

<div style="float: left; width: 50%;">

```{r}
# Basic bar graph
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures)) +
  geom_point() +
  theme_classic()

```

</div>

<div style="float: left; width: 50%;">

```{r}
# Basic scatter plot with cutom color
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures)) +
  geom_point(color = "light blue") +
  theme_classic()

```

</div>

+ When color is set in the `geom_xxx()` it changes color

  - when color is set in `aes()` it looks for a variable


What about using "_decorational_" color schemes?
=========================================

+ Need to add `fill = ethnicity` and `scale_fill_discrete_xxx()` to add color based on `fill`

<div style="float: left; width: 50%;">

+ Using qualitative colors is technically correct

```{r}
# Multi-hued decorative color schemes
dis %>% 
  ggplot(mapping = aes(x = ethnicity, fill = ethnicity)) +
  geom_bar() +
  coord_flip() +
  theme_classic() +
  scale_fill_discrete_qualitative(guide = FALSE)

```

</div>

<div style="float: left; width: 50%;">

+ But this sequential gradient is less cluttered

```{r}
# Single-hued decorative color schemes
dis %>% 
  ggplot(mapping = aes(x = ethnicity, fill = ethnicity)) +
  geom_bar() +
  coord_flip() +
  theme_classic() +
  scale_fill_discrete_sequential(guide = FALSE)

```

</div>

+ **In both graphs we set `guide = FALSE` because we do not need a legend**


What about in different geoms?
=========================================

+ Similar, but use `scale_color_discrete_xxx()` since ethnicity is discrete

  - **the `scale` function must match the attribute in the geom and variable**
  - for example, if we wanted to specify point color by income we would use `scale_color_continuous_xxx()`   

<div style="float: left; width: 50%;">

```{r}
# Vary color by discrete variable in scatterplot 
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, color = ethnicity)) +
  geom_point() +
  theme_classic() +
  scale_color_discrete_sequential()

```

</div>

<div style="float: left; width: 50%;">

```{r message = FALSE}
dis %>% 
  group_by(age, ethnicity) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = age, y = med_exp, color = ethnicity)) +
  geom_smooth(se = FALSE) +
  theme_classic() +
  scale_color_discrete_sequential()

```

</div>


Draw attention - make one or a set of elements stand out
=========================================

<div style="float: left; width: 50%;">

```{r}
# Make one item stand out in bar graph
dis %>% 
  ggplot(mapping = aes(x = ethnicity, fill = ethnicity)) +
  geom_bar() +
  coord_flip() +
  theme_classic() +
  scale_fill_manual(values = c("gray", "gray", "red", "gray"), guide = FALSE)

```

</div>

<div style="float: left; width: 50%;">

```{r}
# Make one item stand out in clustered bar graph
dis %>% 
  ggplot(mapping = aes(x = ethnicity, fill = interaction(ethnicity, gender))) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_classic() +
  scale_fill_manual(values = c("dark gray", "dark gray", "dark red", "dark gray","light gray", "light gray", "red", "light gray"), guide = FALSE)

```

+ The use of interaction() can be tricky, but allows using `fill` with multiple variables

</div>


Draw attention - make one or a set of elements stand out
=========================================

<div style="float: left; width: 50%;">

```{r}
# Single color scheme with scatter plot
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, color = ethnicity)) +
  geom_point() +
  theme_classic() +
  scale_color_manual(values = c("gray", "gray", "red", "gray"))

```

</div>

<div style="float: left; width: 50%;">

```{r message = FALSE}
# Single color scheme with line graph made from geom_smooth()
dis %>% 
  group_by(age, ethnicity) %>% 
  summarize(med_exp = median(expenditures)) %>% 
  ggplot(mapping = aes(x = age, y = med_exp, color = ethnicity)) +
  geom_smooth(se = FALSE) +
  theme_classic() +
  scale_color_manual(values = c("gray", "gray", "red", "gray"))

```

</div>

+ We will do more with this in annotation

Qualitative
=========================================

<div style="float: left; width: 50%;">

+ The goal of using a qualitative palette is to have no individual value stand out

  - this can be challenging:
    - can look cluttered or discordant
    - set palette might have a color that does stand out

```{r color1, eval = FALSE}
# Qualitative color scheme with bar graph
dis %>% 
  ggplot(mapping = aes(x = ethnicity, fill = gender)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_classic() +
  scale_fill_discrete_qualitative(palette = "Dark 2")

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "color1", echo = FALSE}
# Get previous graph

```

</div>


Sequential - show order of elements 
=========================================

<div style="float: left; width: 50%;">

+ The goal of sequential is to create order

  - but this can take some effort
  - see here we have to use `reorder()` twice to get colors and count in consistent order

+ I feel sequential is easiest on the eye and uncluttered
  
```{r color2, eval = FALSE}
# Sequential color scheme with bar graph
dis %>% 
  group_by(ethnicity) %>% 
  summarise(count_n = n()) %>% 
  ggplot(mapping = aes(x = reorder(ethnicity, count_n), y = count_n, fill = reorder(ethnicity, count_n))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_classic() +
  scale_fill_discrete_sequential(palette = "Purples 2", guide = FALSE)

```

</div>

<div style="float: left; width: 50%;">

```{r ref.label = "color2", echo = FALSE}
# Get previous graph

```

</div>


Diverging - where elements diverge from neutral to two extremes
=========================================

<div style="float: left; width: 50%;">

```{r}
# Diverging color scheme with bar graph
dis %>% 
  ggplot(mapping = aes(x = ethnicity, fill = ethnicity)) +
  geom_bar() +
  coord_flip() +
  theme_classic() +
  scale_fill_discrete_diverging(palette = "Blue-Red 2", guide = FALSE)

```

</div>

<div style="float: left; width: 50%;">

```{r}
# Diverging color scheme with scatter plot
dis %>% 
  ggplot(mapping = aes(x = age, y = expenditures, color = age_group)) +
  geom_point() +
  theme_classic() +
  scale_color_discrete_diverging(palette = "Green-Orange", guide = FALSE)

```

</div>


Wrap up
=========================================

<div style="float: right; width: 95%;">

+ A lot can be done to customize colors in a plot

  - but do not just add color because you can

+ Consider using custom font colors to emphasize and deemphasize

  - labels in gray rather than black to deemphasize
  - labels in red to make stand out

+ However, must use colors carefully 

  - it can be a fine line between emphasizing a finding and creating distraction and clutter

</div>


